---
title: "R Training Day 03"
author: "Abbas Rizvi"
date: "6/19/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Welcome to day 03 of R training! I hope that you were able to learn about some fundamental tidyverse packages last class. Data exploration is foundational to data science. Pre-processing data has an inherent relationship with data exploration. Today we are going to go over practical skills like string manipulation, working with dates, writing functions, and iterating over functions using tidyverse and tidyverse affiliated packages. Working with text obviously has its benefits in other areas of data science too, particularly in topics and research related to natural language processing.

# String manipulation using stringr
The dataset that we will use is the `fruit` dataset that comes internal with the stringr package. The stringr package does not load with tidyverse -- so you need to load them separately.

```{r}
library(tidyverse)
library(stringr)
```

Fruit is a character vector length 80 

```{r}
fruit
fruit %>%
    length()
```

## Detect matches

Specific patterns can be detected using `str_detect`.

```{r}
str_detect(fruit, "a")
``` 

Vectors can be subset using the `[` and `]` operators.

```{r}
fruit[str_detect(fruit, "a")]
``` 

The number of matches in a string can be counted

```{r}
str_count(head(fruit), "p")
```

The positions of the pattern in a string can be found as well.

```{r}
str_locate(head(fruit), "p")
str_locate_all(head(fruit), "p")
```

## Subset strings
Substrings can be extracted from character vectors

```{r}
str_sub(fruit, 1,3)
```

As before with the vector subsetting, below is another way to return only strings that contain a pattern match.

```{r}
str_subset(head(fruit), "a")
```

A very useful function is `str_extract`. 

```{r}
str_extract(head(fruit), "a")
str_extract_all(head(fruit), "a")
```

Notice the difference? `str_extract_all` returns every pattern match as opposed to just the first. `str_extract` returns only the first pattern match found in each string, as a vector.

## Manage lengths

Determining the length of a string is a powerful tool, especially in NLP.

```{r}
str_length(fruit)
```

Sometimes (for example in that WHO dataset), we want to pad numbers say that they make more sense organizationally. For example

Computers will order 002, 022, 222, in the proper order but necessarily 2, 22, 222; so a user may want to pad those numbers. The str_pad command permits that.

```{r}
str_pad(string=c(2, 22, 222), 
        side="left",
        pad="0", 
        width=3)
```

Sometimes, you will receive data that has a column full of text and there are random white spaces. Users can remove those using the `str_trim()` function.

```{r}
str_trim("    this is a weird sentence   ", 
         side="both")
```

## Mutate strings

We have already gone over `str_sub`, however, there are several ways to mutate strings using stringr. Let's look at the dataset `sentences` in `stringr`.

```{r}
data(sentences)
# lower case all the sentences
sentences <- sentences %>%
    str_to_lower()

# replace the first 'but' with 'REPLACED'
sentences %>% 
    str_subset("but") %>%
    head() %>%
    str_replace("but", "REPLACED")

# replace all of the 'but' with 'REPLACED'
sentences %>% 
    str_subset("but") %>%
    head() %>%
    str_replace_all("but", "REPLACED")

# make the string into title format (capitalized first letter)
sentences %>%
    head() %>%
    str_to_title
```

## Join and split
Sometimes users will want to either join or split you strings. 

If a user wants to join multiple strings together into a single string.

```{r}
str_c(letters, LETTERS)
```


```{r}
str_dup(head(fruit), 2)
```

A user may want to split by a space or an underscore (or perhaps a more complicated pattern).

```{r}
str_split(head(fruit),
          pattern=" ")
```

Notice that str_split returns a list of substrings that are broken by the delimiter that was specified.

Unfortunately, this is all we will cover for stringr, However, there are many other functions in stringr that should be explored! It is a wonderful package. Explore https://stringr.tidyverse.org
 
## Regular Expressions (regex)
Regular expressions (regex) are a concise language for describing patterns in strings. Often it is said, when you have a regex problem, you have now have two problems, 1. the original problem, and now, 2. regex. 

While I'm sure some of you will become regex masters, it is totally normal to have to look up regular expression syntax when working on your problem.

# Working with dates and lubridate
As working with dates a very important part of data analysis and in particular, time series analysis, we thought it would be worthwhile to introduce the package lubridate.

Date-time data can sometimes be frustrating to worth with. Base R commands for date-times are not necessarily the most intuitive and are not consistent. Lubridate takes is quite detailed and robust to time zones, leap days, day light savings times, and other time related nuances. Lubridate makes these relatively approachable. 

The lubridate package is not included in the tidyverse core packages and thus has to be loaded separately. 

```{r}
library(tidyverse)
library(lubridate)
```

For lubridate we will use the `lubridate::lakers` dataset

```{r}
data(lakers)
lakers <- as_tibble(lakers)
```

# Convert date to date-time

Let's convert the date to proper date object. We will use `lubridate::ymd`.
```{r}
(
    lakers <- lakers %>%
        mutate(date=ymd(date)) 
)
```
Notice has the date column shows `<date>` as the class for that column. 

We can make a new column to add the time. A few pre-processing steps are needed for using some of the packages we saw before.

```{r}
(
    lakers <- 
        lakers %>% 
        unite(date_time, c(date, time), sep=" ") %>%
        mutate(date_time=ymd_hm(date_time))
)
```

```{r}
range(lakers$date_time)
```

If the dates were in month day year (`mdy()`) or quarterly (`yq()`), lubridate has functions for those types of formats as well. 

# Extract month, day, year

```{r}
lakers %>%
    select(date_time) %>%
    mutate(month=month(date_time),
           day=day(date_time),
           wday=wday(date_time),
           wday_label=wday(date_time, label=TRUE),
           year=year(date_time))
```

There are many other functionalities of lubridate, but hopefully this will get you excited about its capabilities. It also can deal with time zones, date-time math, intervals, periods, and duration. Check out https://lubridate.tidyverse.org for more details.

# Iteration

Iteration means repetition of a process. Here we are going to show how we can automate tasks through iteration. The purrr package in tidyverse will be used here. But first we need to discuss lists. 


## Lists
Lists are an object in R which contain elements of different types (like numbers, strings, vectors). Lists can also contain matrices or functions. A list can be created using the `list()` function.


```{r}
?list
```

Here is a list

```{r}
a_list <-
    list(
        num=c(8,9),
        log=TRUE,
        cha=c("a", "b", "c")
    )
```

Shown below are two ways to subset. Do they return the same values?
```{r}
a_list["num"]
a_list$num
```

No they don't. 

```{r}
a_list["num"]
```

is a list with one lement named "num" that contains an atomic vector

```{r}
a_list$num
```
is an atomic vector

```{r}
a_list[["num"]]
```

Lists can be a useful way to organize data.

But you need to manually arrange for functions to iterate over the elements of a list. 

## Creating toy dataset

Here we will simulate some data.

```{r}
set.seed(333)
students_grades <- list(
    apollo = round(runif(10, 50, 100)),
    jimmy = round(runif(10, 50, 100)),
    barbara = round(runif(10, 50, 100)),
    jessica = round(runif(10, 50, 100)),
    fatima = round(runif(10, 50, 100))
)
```

How can we compute the average?

```{r}
mean(students_grades)
```

Instead of doing it brute force like 

```{r}
list(apollo=mean(students_grades$apollo),
     jimmy=mean(students_grades$jimmy),
     barbara=mean(students_grades$barbara),
     jessica=mean(students_grades$jessica),
     fatima=mean(students_grades$fatima))
```

## purrr
We could use purrr.

```{r}
students_grades %>%
    map(mean)
```

The map function applies a function every element of a list. And it returns the results a list. 

`map(.x, .f, ...)` where:   
  * `.x` = a list  
  * `.f` = a function to apply to each element of the list (element argument become first argument)  
  * `...` = other arguments to pass to the function 
  
 There are many more map functions:
 | function  | returns results as      |
|-----------|-------------------------|
| map()     | list                    |
| map_chr() | character vector        |
| map_dbl() | double vector (numeric) |
| map_int() | integer vector          |
| map_lgl() | logical vector          |
| map_df()  | data frame              |

So if we wanted to output the last result as a vector we could use `map_dbl()`

```{r}
students_grades %>%
    map_dbl(mean)
```

The additional arguments could be for something like the `prob` function.

```{r}
students_grades %>%
    map_dbl(quantile, prob=0.9)
```

One of my favorite cases for using purrr is when I want to read in multiple files and bind them all together using `map_dfr()`. 
```{r, eval=FALSE}
files <- dir(".csv") # look for files that end in .csv
files %>%
    map_dfr(read_csv)
```

## Writing functions in R
Naturally, purrr is highly useful when writing your own functions. Here we will go over the basics of writing functions in R. We will walk through this step by step. I'll often write a function to pre-process a bunch of files as I'm reading them in, so I don't read in data that I am not currently interested in using up my resources (i.e. RAM).

Imagine there is a task you would like to do many times:  
  1. Write the code that solves the problem.   
  2. Wrap the code in `function(){} ` to save it  
  3. Generalize your code with variable names  
  4. Start calling your function  

For example, what if we wanted to compute the average again for those students, but this time, we drop the lowest (minimum) grade.

You can write a function:

```{r}
adjusted_grade <- function(current_grades){
    (sum(current_grades) - min(current_grades)) / (length(current_grades) - 1)
}
```

We can test this on one of the students
```{r}
adjusted_grade(students_grades[["apollo"]])
```

Now we can use purrr and apply to all of them.

```{r}
students_grades %>%
    map_dbl(adjusted_grade)
```

Users could also write the function directly into the map function.

```{r}
students_grades %>%
    map_dbl(function(current_grades)  (sum(current_grades) - min(current_grades)) / (length(current_grades) - 1))
```

## map2
There is also map2, which applies a function to every elemnt of two lists. And returns the results a list.  

`map2(.x, .y, .f, ...)`   
  * `.x` = a list of elmeents to pass to the first argument of `.f`   
  * `.y` = a list of elements to pass to the second argument of` .f`  

Suppose we now want to add extra credit for each of the students.

```{r}
extra_credit <- list(10, 12, 35, 10, 15)
```

The grades with extra credit
```{r}
add_extra_credit <- function(current_grades, extra_credit){
    sum(current_grades + extra_credit) / length(current_grades)
}

students_grades %>%
    map2_dbl(extra_credit, add_extra_credit)

```

So we can see, especially for Barbara ... DEFINITELY do your extra credit!

So that was a glimpse at purrr. There are other purrr functions that are great for iteration. Check them out at https://purrr.tidyverse.com







