---
title: "Day 01 - Introduction to R"
date: 'June 15, 2020'
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Get a little taste of R

Take a look at data

```{r}
## I already installed ggplot2 package for you
## So you just need to load it
library(ggplot2)

## We will use the economics data from ggplot2
head(economics)
```

Type `?economics` in console to see column descriptions.

Let's plot total US population in thousands from 1967 to 2015.

```{r}
ggplot(economics, aes(date, pce/pop)) +
  geom_line()
```

Now, fill in the blanks in the code below to plot _personal consumption expenditures_, in billions of dollars.

```{r, eval=FALSE}
ggplot(economics, aes(date, ___)) +
  geom_line()
```


What if we wanted to adjust `pce` by population?
Fill in the blanks.


```{r, eval=FALSE}
ggplot(economics, aes(date, ___/___)) +
  geom_line()
```


## Running code lines

Two options:

1. Go to the line you would like to run and press *Cmd+Enter* (*Ctrl+Enter* on Windows)    
2. Or go to the line you would like to run and click on "Run" on top right.

For example

```{r}
ggplot(economics, aes(date, unemploy))+
         geom_line()

ggplot(economics, aes(date, psavert)) +
  geom_line()
```

 
# R Objects

## Assigning R objects

Save information as an R object with the greater than sign followed by a minus, e.g, an arrow: `<-`

```{r}
foo <- 42
foo
```

When you create an R object, you'll see it appear in your **Environment pane**

## Common R workflow

Usually in R you would assign the output of a function to a variable to save it and then to use it in a second function.

```{r}
foo <- round(3.1415) + 1 
foo
factorial(foo)
```

Because if you don't save the output in a variable, you would only output it to the console. And can't use it as an input for the next step

```{r}
round(3.1415) + 1 
```

## Object names

 1. Object names cannot begin with numbers, or R special characters.
 These as variable names will give you errors: `1st`, `$income`, `>data`, `!bad`, `^mean` 
 
 2. Be careful when reassigning names already in use.
 
 3. Capitalization matters. R will treat each of these as a different object:
 
 `MyData` vs `mydata`
 `A` vs `a`
 `sum` (a R function) vs `SUM`
 
 4. You can remove an object (including the data contained in it) with `rm` function.
 Simply put the name of the variable in the `rm` e.g. `rm(foo)`
 

# Data classes

Like Excel, R can recognize different types of data.
We'll look at four basic types:

 * numbers     
 * character strings (text)   
 * logical   
 * factor
 
Check the type (class) of any R object with `class()` function 
 
## Numeric

Any number, no quotes.
Appropriate for math.
 
```{r}
1+ 1
3000000
class(0.00001)
```

## Character

Any symbols surrounded by quotes.
Appropriate for words, variable names, messages, any text.

```{r}
"hello"
class("hello")
```

Check the number of characters in a character element.
```{r}
nchar("hello")
```

Merge two character elements into one with `paste()`

```{r}
paste("hello", "world")
```

## Logical

`TRUE` (or just `T`) or `FALSE` (or just `F`)
R's form of binary data. Useful for logical tests.

```{r}
3< 4
class(TRUE)
class(T)
class(FALSE)
class(F)
```

## Factor

R's form of categorical data.
Especially useful for modeling and plotting.
Saved as an integer with a set of labels (e.g. levels).

```{r}
vec2 <- c("a", "b","b", "c", "d", "a") # vec2 is a character vector
fac <- factor(vec2) #now fac is factor
class(fac)
fac
```

One can easily change the "labels" e.g. `levels` of a factor

Let's change `a` to `apple` and `d` to `dragonfruit`


```{r}
fac
levels(fac)
levels(fac) <- c("apple", "b", "c", "dragonfruit")
fac
```

# Data Structures

## Vectors
Concatenate multiple elements into a one dimensional array.
Create with the `c()` function.

```{r}
vec <- c(1, 2, 3, 10, 100)
vec
```

## Matrices

Multiple elements stored in a two dimensional array.
Create with the `matrix()` function

```{r}
mat <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2)
mat
```

You can also transpose a matrix with `t()` function.

```{r}
t(mat)
```


## Math: element-wise

```{r}
vec
vec + 4

vec * 4

vec * vec
```


Vector and matrices cannot have different class elements. For example a vector cannot contain a character, a number or a logical all at the same time. Therefore if there are **any** characters, all other elements of the vector are coerced to character class. 

```{r}
v1 <- c("R", 1, TRUE)
class(v1)
v1
```

If the vector contains numeric and logical elements, all will coerce to numeric, where `TRUE` equals to `1` and `FALSE` equals to `0`.

```{r}
v2 <- c(1, TRUE)
class(v2)
v2
```

**What if you want different data types in the same object?**

List and data frames generalize vectors and matrices to allow multiple types of data.

## List
A list is a one dimensional group of R objects.
Create lists with `list()`

```{r}
lst <- list(1, "R", TRUE) 
class(lst)
lst
```

The elements of a list can be anything. Even vectors, matrices, data frames or other lists!

```{r}
list(c(1, 2), TRUE, c("a", "b", "c"), matrix(c(1:10), nrow = 2))
```

## Data frame

The classic data table we see day to day...
A data frame is a two dimensional group of R objects.
Each column in a data frame can be a different type (class).
(But the elements of a column cannot be of different type.)

```{r}
df <- data.frame( c(1, 2, 3),  #this one is numeric
                  c("R","S","T"), # this one is character
                  c(TRUE, FALSE, TRUE)) #this one is logical
class(df)
df
```

## Names

You can name the elements of a vector, list, or data frame when you create them.

```{r}
#vector
nvec <- c(one = 1, two = 2, three = 3)
nvec

# list
nlst <- list(one = 1, 
             two = 2,
             many = c(3, 4, 5))
 
nlst

# data frame
ndf <- data.frame(numbers = c(1, 2, 3),
                  letters = c("R","S","T"),
                  logic = c(TRUE, FALSE, TRUE))
```

You can also see and set the names with `names()`. (Similar to `levels()`, do you remember that one?)

```{r}
names(nvec)
names(nvec) <- c("uno", "dos", "tres") 
nvec
```

Similarly you can change the name of rows and columns of a data frame as well.

```{r}
colnames(ndf)
colnames(ndf) <- c("nombres", "lettres", "logique") # yes, it's french
ndf

rownames(ndf)
rownames(ndf) <- c("myRow1", "myRow2", "myRow3")
ndf
```



Once you have the names set up, you can call columns of a data frame or elements of a vector or a list very easily.

```{r}
# vector
nvec["uno"]

# list -- attention to double brackets
nlst[["many"]]
# or another way
nlst$many

# data frame
ndf["nombres"]
# or another way
ndf$nombres
```

Can you tell the difference between `ndf["numbers"]` and `ndf$numbers`?   
They're different type of data structures!    
`ndf["nombres"]` is a data frame with only one column, `ndf$nombres` is a numeric vector.

```{r}
class(ndf["nombres"])
class(ndf$nombres)
```

# Let's not forget

 1. There are data classes (types) such as: 
  * character 
  * numeric 
  * logical 
  * factor
 2. There are also data structures: 
  * vector
  * matrix
  * list
  * data frame
 3. Some data structures **can't** contain different classes (types) of data (e.g. they can only contain one type of data):
  * vector
  * matrix
4. Some data structures **can** contain different classes (types) of data (e.g. they can contain multiple types of data):
  * list
  * data frame


